A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-05-02 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: reworked the introduction

>>>>

In the past few decades, there has been a growing interest in developing approaches that generalize the notion of states and physical theories. One basic insight is that statistical/probabilistic concepts often provide a suitable basis for this generalization, and it is in this setting where most of these approaches operate. A key problem, then, is what notion of probability should be used given that the standard measure theoretic Kolmogorovian approach does not generally work in quantum mechanics.

We reformulate the problem in the following way: given the space of all possible statistical ensembles allowed by a statistical theory, what measure theoretic tools can we use to represent each ensemble? If they are not always available, in what circumstances and with what limitations are they available? Is a suitable generalization possible, and what mathematical tools would need to be developed?

In this paper, we will define a space of ensembles as a convex bounded set of a Hausdorff locally convex topological vector space, upon which an entropy function is defined. These mathematical properties are physically grounded in the ability to experimentally define ensemble (leading to a topological space), to define statistical mixtures (leading to a convex space), to quantify the variability of the elements of the ensembles (leading to an entropy functions) and to having enough real valued quantities to identify each ensemble (leading to the continuous embedding into a Hausdorff locally convex topological vector space). The question is, then, what type of measure theoretic tools can we use to quantify the number of states and to represent each ensemble in the space. Both are needed to be able to define a physically meaningful notion of probability density.

We will first review some results from Choquet theory, which is, to our knowledge, the most general approach to represent elements of a convex set using additive probability measures. The key insight is that, at least in the compact case, such representation is always possible, but it is not unique. Uniqueness requires the space to be a Choquet simplex, which is a generalization of the standard finite-dimensional simplex. Quantum theory does not fall in this case: it is precisely the multiple decomposition of mixed states into pure states that prevents a quantum ensemble space to be simplex.

We will also review some results from the use of signed probability measures. While the most known example is the use of Wigner functions, there are actually several ones and there is also work on developing a general theory of signed probability. Given that the space of signed probability measure is, unlike a simplex, a vector space, we can find an affine map between signed probability measures and the embedding vector space. This gives us a unique representation of each ensemble in term of a singed probability measure. However, unlike in the Choquet case, the measure is not supported by the set of extreme points of the space. That is, there is no unique or privileged choice for this representation. Moreover, since the map is at the level of the full vector spaces, there is no general way to characterize which subset of signed probability measures correspond to ensembles.

At this point, we will show that we can define two types of measures: the state capacity, which generalizes the notion of count of states, and a fraction capacity for each ensemble, which generalizes the notion of probability. Both are non-negative sub-additive monotonic set functions, which the second being unit bounded. These are defined on the ensemble space itself, which means the domain is uniquely defined by the ensemble space. Moreover, in the compact case, these measure can be restricted on the extreme points and still retain uniqueness. The fraction capacity, in fact, reduce to the supremum of all Choquet measures over the extreme points. The classical case exactly corresponds to the simplex case, in which fraction capacity corresponds to the unique measure, and is therefore additive. The state capacity, instead, recovers the Liouville measure in the classical case and the dimensionality of the Hilbert space in the quantum case.

Having shown that a general approach based on non-additive measures is possible, we lay down a series of questions to explore in the future. The biggest one is whether it is possible to generalize notions such as integration, derivation and expectation to this non-additive case.  It should be noted that non-additive measures are used in the context of decision theory, so there may be a connection between the use of non-additive measures in quantum theory and interdependent complimentary choices.[TODO: find citation that sub-additive is when choices are complementary] Both the Choquet and the Sugeno integral, for example, are defined over a monotonic set-functions, so there is a chance that these tools may turn out to be useful. We see this work as starting this line of research, which will take more effort to investigate.

\section{Old introduction}
